### 2023-07-20 doc/_bayesian_inference.qmd
making progress on writing the paper

---
editor:
    render-on-save: true
---
## Bayesian inference {#sec-bayesian_inference}

As defined in @sec-fitness_model, our ultimate objective is to infer the vector
of relative fitness values
$$
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
$${#eq-fitness_mut_vec}
where $^\dagger$ indicates the transpose. Our data consists of an $B \times T$
matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
barcodes given by the sum of the number of unique mutant barcodes $M$ and the
number of unique neutral barcodes $N$, and $T$ is the number of time points
where measurements were taken. The data matrix is then of the form
$$
\underline{\underline{R}} = \begin{bmatrix}
\lvert & \lvert & \cdots & \lvert\\
\underline{r}_1 & \underline{r}_2 & \cdots & \underline{r}_T\\
\lvert & \lvert & \cdots & \lvert
\end{bmatrix},
$${#eq-R_mat}
where each column $\underline{r}_t$ is a $B$-dimensional array containing the
raw barcode counts at time $t$. We can further split each vector
$\underline{r}_t$ into two vectors of the form
$$
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{M} \\
\underline{r}_t^{N}
\end{bmatrix},
$${#eq-r-vec_split}
i.e., the vector containing the mutant barcode counts $\underline{r}_t^{M}$ and
the corresponding vector containing the neutral barcode counts
$\underline{r}_t^{N}$. Our task is to compute the joint probability distribution
of all relative fitness values given our data. Writing Bayes theorem this is
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines the probability distribution  of the input
unless otherwise stated. We can ignore the denominator on the right-hand side
of @eq-bayes_obj, since it serves as a normalization constant that does not
depend on our parameter of interest.

Although @eq-bayes_obj seems simple enough, recall that @eq-fitness relates
frequency values and the population mean fitness to the mutant relative fitness.
Therefore, we must include these nuisance parameters as part of our inference
problem. At the end of the day, we can always marginalize the joint distribution
to obtain the distribution we are interested in. To include these parameters,
let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires time points $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $B \times T$ matrix containing all
frequency values. As with @eq-R_mat, each column of $\underline{\underline{F}}$
can be split into two components of the form
$$
\underline{f}_t = \begin{bmatrix}
\underline{f}_t^M \\
\underline{f}_t^N
\end{bmatrix},
$${#eq-f_vec}
to split the corresponding mutant and neutral barcode frequencies. With these
variables in hand, the full inference problem we must solve takes the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
As a reminder, we need to include these extra nuisance parameters, but we can
always integrate them out to recover the marginal distribution we care about,
i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}
Although daunting at first sight, @eq-bayes_full_marginal will be solved 
numerically, so we do not have to worry about this integration.

### Splitting the posterior distribution {#seq-split_posterior}

As written, the left-hand side of @eq-bayes_full is really difficult to work
with. However, we can take advantage of the structure of our inference problem
to rewrite it in a more manageable form. Specifically, we rewrite the joint
probability distribution as a product of conditional distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}},
    \underline{\underline{R}}
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}, 
    \underline{\underline{R}}
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of 
uncertainty listed in @sec-fitness_model in each of the terms. Starting from
right to left, the first term on the right-hand side of @eq-split_posterior
accounts for the uncertainty when inferring the frequency values given the
barcode reads. The second term accounts for the uncertainty in the values of
the mean population fitness at different time points. The last term accounts for
the uncertainty in the parameter we care about---the mutants' relative
fitnesses. In the following sections, we work through each of these terms to
build up the posterior distribution we will sample from piece by piece.

>>>>

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---
## Bayesian inference {#sec-bayesian_inference}

As defined in @sec-fitness_model, our ultimate objective is to infer the vector
of relative fitness values
$$
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
$${#eq-fitness_mut_vec}
where $^\dagger$ indicates the transpose. Our data consists of an $T \times B$
matrix $\underline{\underline{R}}$, where $B = M + N$ is the number of unique
barcodes given by the sum of the number of unique mutant barcodes $M$ and the
number of unique neutral barcodes $N$, and $T$ is the number of time points
where measurements were taken. The data matrix is then of the form
$$
\underline{\underline{R}} = \begin{bmatrix}
- & \underline{r}_1 & - \\
- & \underline{r}_2 & - \\
 & \vdots & \\
- & \underline{r}_T & - \\
\end{bmatrix},
$${#eq-R_mat}
where each row $\underline{r}_t$ is a $B$-dimensional array containing the
raw barcode counts at time $t$. We can further split each vector
$\underline{r}_t$ into two vectors of the form
$$
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{N} \;
\underline{r}_t^{M}
\end{bmatrix},
$${#eq-r-vec_split}
i.e., the vector containing the neutral lineages barcode counts
$\underline{r}_t^{N}$ and the corresponding vector containing the mutant barcode
counts $\underline{r}_t^{M}$. Our objective is to compute the joint probability
distribution of all relative fitness values given our data. We can express this 
joint distribution using Bayes theorem as
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
$${#eq-bayes_obj}
where hereafter $\pi(\cdot)$ defines a probability density function unless
otherwise stated. We can ignore the denominator on the right-hand side of
@eq-bayes_obj, since it serves as a normalization constant that does not depend
on our parameter of interest, simply writing
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)
$${#eq-bayes_obj_propto}

Although @eq-bayes_obj and @eq-bayes_obj_propto seem simple enough, recall that
@eq-fitness relates barcode frequency values and the population mean fitness to
the mutant relative fitness. Therefore, we must include these nuisance
parameters as part of our inference problem. We can always marginalize the joint
distribution to obtain the distribution we are interested in. To include these
parameters, let 
$$
\underline{\bar{s}}_T = (\bar{s}_1, \bar{s}_2, \ldots, \bar{s}_{T-1})^\dagger,
$${#eq-pop_fitness_vec}
be the vector containing the $T-1$ population mean fitness we compute from the
$T$ time points where measurements were taken. We have $T-1$ since the value
of any $\bar{s}_t$ requires time points $t$ and $t+1$. Furthermore, let the
matrix $\underline{\underline{F}}$ be a $T \times B$ matrix containing all
frequency values. As with @eq-R_mat, each row of $\underline{\underline{F}}$
can be split into two components of the form
$$
\underline{f}_t = \begin{bmatrix}
\underline{f}_t^M \;
\underline{f}_t^N
\end{bmatrix},
$${#eq-f_vec}
to separate the corresponding mutant and neutral barcode frequencies. With these
variables in hand, the full inference problem we must solve takes the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
$${#eq-bayes_full}
We can numerically integrate out all nuisance parameters to recover the marginal
distribution we care about, i.e.,
$$
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
$${#eq-bayes_full_marginal}

### Factorizing the posterior distribution {#seq-split_posterior}

The left-hand side of @eq-bayes_full is extremely difficult to work with.
However, we can take advantage of the structure of our inference problem to
rewrite it in a more manageable form. Specifically, we rewrite the joint
probability distribution as a product of conditional distributions of the form
$$
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}},
    \underline{\underline{R}}
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}, 
    \underline{\underline{R}}
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
$${#eq-split_posterior}
Written in this form, @eq-split_posterior captures the three sources of 
uncertainty listed in @sec-fitness_model in each of the terms. Starting from
right to left, the first term on the right-hand side of @eq-split_posterior
accounts for the uncertainty when inferring the frequency values given the
barcode reads. The second term accounts for the uncertainty in the values of
the mean population fitness at different time points. The last term accounts for
the uncertainty in the parameter we care about---the mutants' relative
fitnesses. We invite the reader to check the supplementary material where we
work through each of these terms to build the posterior distribution to sample
from.

### 2023-07-20 doc/_experiment.qmd
making progress on writing the paper

@@

`[add details on expeirmental setup]`

>>>>

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

@@

Our inference pipeline is designed for the type of competition assays presented
in @kinsler2020. In these experiments, an ensemble of genotypes is pooled
together with an unlabeled reference strain that, initially, represents the vast
majority ($\geq 90\%$) of the cells in the culture. Furthermore, a subfraction
of labeled genotypes equivalent to the unlabeled reference strain---hereafter
defined as *neutral* lineages---are spiked in at a relatively high abundance
($\approx 3-5\%$), the rest of the culture being left for the ensemble of
genotypes of interest.

To determine the relative fitness of the ensemble of genotypes, a series of
growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time to then
dilute the culture back into fresh media and repeat for roughly 4-6 cycles,
depending on the initial abundances of the mutants, and their relative growth
rates. The DNA barcodes are sequenced at the end of each growth cycle to
quantify the relative abundance of each of the barcodes. We point the reader to
@kinsler2020 for specific details on these assays for *S. cerevisiae*. Figure
[`ref here`] presents a typical barcode trajectory where the darker barcode
trajectories represent the neutral lineages.

### 2023-07-20 doc/_fitness_model.qmd
making progress on writing the paper

---
editor:
    render-on-save: true
---

## Preliminaries on mathematical notation
Before jumping directly into what concerns us here, let us establish the
mathematical notation to be used throughout this document. We define (column)
vectors as underlined lowercase symbols such as
$$
\underline{x} = \begin{bmatrix}
    x_1\\
    x_2\\
    \vdots\\
    x_N
\end{bmatrix}.
$${#eq-vec_def}

@@

Empirically, we have seen that the barcode trajectories follow an exponential
function of the form
$$
f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
$${#eq-fitness}
where $f_{t}^{(b)}$ is the frequency of barcode $b$ at time $t$, $s^{(b)}$ is
the relative fitness with respect to the ancestral strain---the quantity we want
to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
$t$, and $\tau$ is the time pass between time $t$ and $t+1$. It is important to
notice that the frequencies $f_{t}^{(b)}$ are not the allele frequencies in the
population (most of the culture is not sequenced since it is not barcoded), but
rather the inferred frequencies from the total number of reads. A way to 
conceptualize this is that we assume exponential growth in the *number of cells*
$N_t^{(b)}$ of the form
$$
N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
$${#eq-ncells}
for every barcode $b$. However, when we sequence barcodes, we do not directly
measure the number of cells, but some number of reads $r_t^{(b)}$ that map to
barcode $b$. In the simplest possible scenario, we assume
$$
r_{t}^{(b)} \propto N_{t}^{(b)},
$${#eq-r_to_n}
where, importantly, the proportionality constant depends on the total number of
reads for the time $t$ library, which might vary from library to library.
Therefore, to compare the number of reads between libraries at different time
points, we must somehow normalize them to the same scale. The simplest form is
to define a relative abundance, i.e., a frequency with respect to the total
number of reads,
$$
f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
$${#eq-r_to_n_norm}
This is the frequency @eq-fitness describes.

Our ultimate objective is to infer the relative fitness $s^{(m)}$ for each of
the $M$ mutants in the experiment. To do so, we must take into account the three
main sources of uncertainty in our model:

1. Uncertainty in the determination of frequencies. Our model relates
frequencies between adjacent time points to the fitness of the corresponding
strain. However, we do not measure frequencies directly. Instead, our data
consists of a vector of barcode counts $\underline{r}_t$ for each of the $T$
time points in which the measurements were taken.
2. Uncertainty in the value of the population mean fitness. As we will see later
on, we use neutral lineages barcodes, which, by definition have mean fitness
$s^{(n)}$ to determine the value of the population mean fitness $\bar{s}_t$.
This inference will come with its own uncertainty that must be propagated to the
value of the relative fitness for the mutants.
3. Uncertainty in each of the mutant's fitness values.

>>>>

---
editor:
    render-on-save: true
bibliography: references.bib
csl: ieee.csl
---

## Preliminaries on mathematical notation
Before jumping directly into the Bayesian inference pipeline, let us establish
the mathematical notation to be used throughout this paper. We define
(column) vectors as underlined lowercase symbols such as
$$
\underline{x} = \begin{bmatrix}
    x_1\\
    x_2\\
    \vdots\\
    x_N
\end{bmatrix}.
$${#eq-vec_def}

@@

Empirically, the barcode relative frequency trajectories follow an exponential
function of the form
$$
f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
$${#eq-fitness}
where $f_{t}^{(b)}$ is the frequency of barcode $b$ at time $t$, $s^{(b)}$ is
the relative fitness with respect to the reference strain---the quantity we want
to infer from the data---$\bar{s}_t$ is the mean fitness of the culture at time
$t$, and $\tau$ is the time pass between time $t$ and $t+1$. We can rewrite
@eq-fitness as
$$
\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
$${#eq-logfreq}
@eq-logfreq separates the observables---the barcode frequencies---from the
latent parameters we want to infer from the data---the population mean fitness
and the barcode relative fitness. This is ultimately the functional form used in
our inference pipeline. Therefore, the relative fitness is computed by knowing
the barcode log frequency ratio of each barcode throughout the growth-dilution
cycles.

It is important to notice that the frequencies $f_{t}^{(b)}$ are not the allele
frequencies in the population (most of the culture is not sequenced since the
reference strain is not barcoded), but rather the relative frequencies in the
total number of sequencing reads. A way to conceptualize this subtle but
important point is to assume exponential growth in the *number of cells*
$N_t^{(b)}$ of the form
$$
N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{\lambda^{(b)}\tau},
$${#eq-ncells}
for every barcode $b$ with growth rate $\lambda^{(b)}$. However, when we
sequence barcodes, we do not directly measure the number of cells, but some
number of reads $r_t^{(b)}$ that map to barcode $b$. In the simplest possible
scenario, we assume
$$
r_{t}^{(b)} \propto N_{t}^{(b)},
$${#eq-r_to_n}
where, importantly, the proportionality constant depends on the total number of
reads for the time $t$ library, which might vary from library to library.
Therefore, to compare the number of reads between libraries at different time
points, we must normalize the number of reads to the same scale. The simplest
form is to define a relative abundance, i.e., a frequency with respect to the
total number of reads,
$$
f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
$${#eq-r_to_n_norm}
This is the frequency @eq-fitness describes.

Our ultimate objective is to infer the relative fitness $s^{(m)}$ for each of
the $M$ mutants in the experiment. To do so, we account the three main sources
of uncertainty in our model:

1. Uncertainty in the determination of frequencies. Our model relates
frequencies between adjacent time points to the fitness of the corresponding
strain. However, we do not directly measure frequencies. Instead, our data for
each barcode consists of a vector of counts $\underline{r}^{(b)}_t$ for each of
the $T$ time points in which the measurements were taken.
2. Uncertainty in the value of the population mean fitness. We define neutral
lineages to have fitness $s^{(n)} = 0$, helping us anchor the value of the
population mean fitness $\bar{s}_t$ for each pair of adjacent time points. 
Moreover, we take this parameter as an empirical observable from the data, 
meaning that we do not have a functional form that relates $\bar{s}_t$ to
$\bar{s}_{t+1}$. Thus, we must infer the $T-1$ values of this population mean 
fitness with their uncertainty that must be propagated to the value of
the mutants' relative fitness.
3. Uncertainty in each of the mutants' fitness values.
